---
title: "Stat 585 - Date and Time"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, "tweaks.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: center, middle
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

# Working with date and time

---

## `lubridate` package

- package for working with dates and times

- defines different classes of time: instants, periods, intervals, durations

- defines converter and accessor functions, enables time calculus

---

## Converter functions and instants in time

- time instants: one (absolute) moment in time, e.g. `now()`, `Jan-1-2000`

- easy-to-use converter functions: 
    - date: `ymd`, `mdy`, `dmy`, ...
    - time: `hm`, `hms`, ...
    - date & time: `ymd_hms`, `mdy_hm`, ...
    
- order of letters determines how strings are parsed

- separators are automatically determined, then assumed to be the same

Example:
```{r message=FALSE}
mdy("03-01-2017")
```

---

## Accessor functions


- accessor functions: `year`, `month`, `week`, `wday`, `mday`, `yday`, `hour`, `minute`, ...

- accessor functions can also be used for setting elements of date and time, e.g. `hour(x) <- 12`

Examples:

```{r}
month(now())
wday(now(), label = TRUE)
```

---

## Intervals and Durations

- Intervals have a *start* and an *end* date/time: absolute difference

- Durations are potentially of relative length (months, leap year, leap second, ...)

```{r}
end_date <- now()
# span is interval, years(1) is duration
span <- end_date - years(1)
span
end_date - days(10)
```


---

## Example: Movies

```{r}
movies <- read.csv("../data/movies.csv")
summary(movies$Date) # factor variable: number of movies in theaters
```

---


## Example: Movies (cont'd)

```{r, fig.height = 3.25}
library(lubridate)
movies$Date <- ymd(movies$Date)
summary(movies$Date) # makes more sense

movies %>% ggplot(aes(x = Date)) + geom_histogram(binwidth=7)
```

---
class:inverse
## Your Turn (10 min)

- load the `movies` data set,
- convert the variable `release` to a date format,
- plot a histogram of the variable.
- Check whether the time between the release of a movie and the date is equal to the number of days in theaters. 
- How should we interpret the result of `as.numeric(months(1))`?

---

## Your Turn - solution

```{r fig.height=4.5}
movies$Release <- ymd(movies$Release)
movies$Days2 <- movies$Date - movies$Release
movies %>% ggplot(aes(x = Release)) + 
  geom_histogram(binwidth = 90)
```

---

## Your Turn - solution (cont'd)

```{r}
str(movies$Release)

as.numeric(months(1))/(60*60*24) # that's the average number of days in a month
```
