---
title: "Stat 585 - Reading fixed width files"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, "tweaks.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## "Awkward" text formats and R

The National Climate Data Center at NOAA publishes information on temperature and precipation across a network of stations in the US.

The Data can be accessed through ftp at ftp://ftp.ncdc.noaa.gov/pub/data/ushcn/v2.5, a code book with a description of the data structure is available at
ftp://ftp.ncdc.noaa.gov/pub/data/ushcn/v2.5/readme.txt

```{r}
temp.all<-readLines("../data/USH00132999.raw.tmax", n = 5)
temp.all
```

---

## From the code book

```
2.2.1 DATA FORMAT

          Variable          Columns      Type
          --------          -------      ----

          ID                 1-11        Integer
          YEAR              13-16        Integer
          VALUE1            17-22        Integer
          DMFLAG1           23-23        Character
          QCFLAG1           24-24        Character
          DSFLAG1           25-25        Character
            .                 .             .
            .                 .             .
            .                 .             .
          VALUE12          116-121       Integer
          DMFLAG12         122-122       Character
          QCFLAG12         123-123       Character
          DSFLAG12         124-124       Character
```

---

# Fixed width format

- Fixed formats can be read with `read.fwf()` or `readr::read_fwf()`

- Need to specify column positions: pretty painful to specify

```{r message=FALSE, warning = FALSE}
library(tidyverse)
temps <- read_fwf("../data/USH00132999.raw.tmax",
  col_positions = fwf_positions(
    start=c( 1,13, rep(16+9*0:11, each=4) + c(1,7,8,9)),
    end  =c(11,16, rep(16+9*0:11, each=4) + c(6,7,8,9))))
names(temps) <- c("Station", "Year",
  paste0(rep(c("Value","DMflag", "QCflag", "DSflag"), 12),
         rep(1:12, each=4)))
```


---

# Fort Dodge temps

```{r warning = FALSE, message = FALSE, fig.height =4}
FortDodge <- temps %>%
  gather(key="Month", value="Temp_Max", starts_with("Value"))
FortDodge$Month <- as.numeric(gsub("Value", "", FortDodge$Month))
FortDodge$Temp_Max <- replace(FortDodge$Temp_Max,
                              FortDodge$Temp_Max == -9999, NA)

FortDodge %>% filter(Temp_Max > -100) %>%
  ggplot(aes(x = Year, Temp_Max/100)) +
  geom_point() + facet_wrap(~Month, nrow=2) + geom_smooth(method="lm")
```

---

# USHCN network

  Parsing the information available in the file `ushcn-v2.5-stations.txt` was part of lab 1. This file is also in fixed width format:

  ```
----------------------------------------
  Variable             Columns    Type
----------------------------------------
COUNTRY CODE             1-2    Character
NETWORK CODE               3    Character
ID PLACEHOLDERS ("00")   4-5    Character
COOP ID                 6-11    Character
LATITUDE               13-20    Real
LONGITUDE              22-30    Real
ELEVATION              33-37    Real
STATE                  39-40    Character
NAME                   42-71    Character
COMPONENT 1 (COOP ID)  73-78    Character
COMPONENT 2 (COOP ID)  80-85    Character
COMPONENT 3 (COOP ID)  87-92    Character
UTC OFFSET             94-95    Integer
-----------------------------------------
  ```
---

## First look at the data

```{r}
stations<-readLines("https://raw.githubusercontent.com/Stat585-at-ISU/materials/master/data/ushcn-v2.5-stations.txt", n = 5)
head(stations)
```

---

# reading the fixed width format

```{r message=FALSE,  warning = FALSE}
stations <- read_fwf("https://raw.githubusercontent.com/Stat585-at-ISU/materials-2019/master/labs/ushcn-v2.5-stations.txt",
                     col_positions = fwf_positions(
                       start=c(1,3,4,6,13,22,33,39,42,73,80,87,94),
                       end  =c(2,3,5,11,20,30,37,40,71,78,85,92,95)))
# check that the formats are correct
stations %>% glimpse()

```

---

# Cleaning the data

We need to get the names in:

```{r}
names(stations) <- c("Country code", "Network code", "ID", "Coop ID",
                     "Latitude", "Longitude", "Elevation", "State",
                     "Name", "Comp1", "Comp2", "Comp3", "UTC offset")
```

Which Iowa stations do we have?

```{r}
stations[stations$State == "IA",]$Name
```

---

Now we plot:

```{r warning = FALSE, message = FALSE, fig.height = 4, fig.width = 8}
us <- map_data("state")
stations <- stations %>% mutate(
  Elevation = as.numeric(Elevation)
)
stations %>% ggplot(aes(x = Longitude, y = Latitude)) +
  geom_path(aes(x = long, y = lat, group = group), data = us) +
  geom_point(aes(size = Elevation, colour = factor(`UTC offset`)), alpha = 0.5) +
  scale_size(range=c(0.5, 3.5)) +
  theme_bw() +
  scale_color_brewer(palette="Paired")
```

There are some stations with wrong(?) UTC offsets
