---
title: "Stat 585 - Databases and R"
author: "Heike Hofmann and Susan Vanderplas"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, "tweaks.css"]
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE, message=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi= 300, fig.width = 5, fig.height = 5)
options(width=60)
library(tidyverse)
```

## Outline

- a bit about databases

- linking R to a (local) data base

- investigate the FARS data

<br><br><br><br>

Resources:

- RStudio's [approach to databases](https://db.rstudio.com/)

- `vignette("two-table", package="dplyr")` (for joining tables - material that was covered in Stat 579)

---

## Why databases?

What is a database?

- A collection of data

- A set of rules to manipulate data

<br><br>
Why are databases important?

- Efficient manipulation of large data sets

- Convenient processing of data

- Integration of multiple sources of data

- Access to a shared resource

---

## Relational Databases

- Database is collection of tables and links (normal form)

- SQL (Structured Query Language) for querying 

- DBMS (Database Management System) and managing data

---

## Structured Query Language (SQL)

- Structured Query Language (1970, E Codds)

- Programming language used for accessing data in a database

-  ANSI standard since 1986, ISO standard since 1987

- Still some portability issues between software systems!

- We'll mainly focus on SQL queries to access data

---

## Syntax

-  SQL is not case sensitive.

- Some systems require a semi-colon `;` at the end of each line. 

The semi-colon can be used to separate each SQL statement in a system that allows multiple command to be executed in a call to the server.


---

## Select

.pull-left[ 
**Student**

| ***ID***            | Name          | Major  |
| ------------- |:-------------:| ------:|
| 1234          | Never Ever    | Math   |
| 4321          | Some Times    | CS     |
]
.pull-right[
**Attendance**

| ***ID***            | ***Date***          | Status  |
| ------------- |:-------------:| -------:|
| 1234          | 02-Feb        | absent  |
| 1234          | 09-Feb        | absent  |
| 4321          | 02-Feb        | present |
| 4321          | 09-Feb        | absent  |
]

<br>

`Select * from student` leads to 

| *ID*           | Name          | Major  |
| ------------- |:-------------:| ------:|
| 1234          | Never Ever    | Math   |
| 4321          | Some Times    | CS     |




---

## Where

.pull-left[
**Student**

| ***ID***            | Name          | Major  |
| ------------- |:-------------:| ------:|
| 1234          | Never Ever    | Math   |
| 4321          | Some Times    | CS     |
]
.pull-right[
**Attendance**

| ***ID***            | ***Date***          | Status  |
| ------------- |:-------------:| -------:|
| 1234          | 02-Feb        | absent  |
| 1234          | 09-Feb        | absent  |
| 4321          | 02-Feb        | present |
| 4321          | 09-Feb        | absent  |
]

<br>

`Select Name from student where Major='Math'` leads to 

| Name          | 
| ------------- |
| Never Ever    | 

---

## Aggregations

.pull-left[
**Student**

| ***ID***            | Name          | Major  |
| ------------- |:-------------:| ------:|
| 1234          | Never Ever    | Math   |
| 4321          | Some Times    | CS     |

]
.pull-right[
**Attendance**

| ***ID***            | ***Date***          | Status  |
| ------------- |:-------------:| -------:|
| 1234          | 02-Feb        | absent  |
| 1234          | 09-Feb        | absent  |
| 4321          | 02-Feb        | present |
| 4321          | 09-Feb        | absent  |
]

<br>
`Select ID, count(ID) from Attendance where Status='absent'` leads to 

| *ID*            | Frequency          | 
| ------------- |-------------:| 
| 1234          | 2    | 
| 4321          | 1    | 

---

## Using SQL in R

package `sqldf` provides a wrapper for sql statements in R:

```
sqldf::sqldf("Select ID, count(ID) from Attendance where Status='absent'")
```

Make sure to use a different set of quotes  on the inside of a statement 

```{r}
sqldf::sqldf("select * from iris limit 5")
```

---
class: inverse
# Your Turn

- Use a SQL command wrapped in`sqldf` to create a table of Species (in the `iris` data)

- Modify the command such that you only consider observations where petal length is less than 4 (Hint: sqldf does not like `.` in variable names - wrap the name in quotes)


---

## Additional Resources for SQL 

- W3Schools: http://www.w3schools.com/sql/default.asp

- DBI package, RMySQL, RSQLite: 
http://cran.r-project.org/web/packages/RMySQL/RMySQL.pdf

http://cran.r-project.org/web/packages/RSQLite/RSQLite.pdf

---

## Connecting to a database

The package `DBI` allows to connect to various different DBMS, such as SQLite, MySQL, PostgreSQL, MonetDB, ...

`dbConnect` and `dbDisconnect`

`dbConnect` needs a DBMS specific driver, drivers have the form `SQLite(), MySQL(), ...`

Connecting information depends on DBMS - e.g. MySQL needs a server, the name of the database,  the user's information and generally a  password

---

## SQLite

- SQLite is a relational database management system

- Unlike other DBMS SQLite is a server-less system with "zero-configuration"


---

## An example: FARS data

- US Department of transportation is keeping a record of every accident that results in a fatality in the FARS Data base (fatal accident report system, http://www.nhtsa.gov/FARS)

- FARS consists of 20+ tables consisting of various aspects of each accident
Documentation at

- http://www-nrd.nhtsa.dot.gov/Cats/listpublications.aspx?Id=J&ShowBy=DocType


---

## FARS 2014

- `fars2014` is a SQLite database of the 2014 data provided by FARS, consisting of three tables: `accidents`, `person`, and `vehicle` (make sure to download  [`fars2014`](https://stat585-at-isu.github.io/materials-2019/data/fars2014)  from our github repo).


connecting to the database:

```{r}
library(dplyr)
fars <- src_sqlite("../data/fars2014")
fars
```

**The life cycle warning is super annoying - we will take care of that later**
---

## Working with sql and dplyr


- `dplyr` package works (almost) the same for local data frames as tables in a database 

- `dplyr` functionality:
`group_by`, `summarize`, `transform`, `filter`, `arrange`, `select`


---



## Connecting to FARS

Use `tbl` to connect to a specific table (dataset) in the database

note: R does not load the data into the session

```{r}
accidents <- tbl(fars, "accidents")
accidents
```

---

##  Collecting results

- `dplyr` commands do not actually download the (whole) database

- use `collect()` to extract *all* records. This returns a tibble

```{r eval = FALSE}
accidents %>% 
  filter(between(LONGITUD, -130, 0)) %>% 
  collect() %>%
  ggplot(aes(LONGITUD, LATITUDE)) +
    geom_point(alpha = 0.5, size = 0.5)
```


---

```{r echo=FALSE, fig.width = 8, fig.height = 6}
accidents %>% 
  filter(between(LONGITUD, -130, 0)) %>% 
  collect() %>%
  ggplot(aes(LONGITUD, LATITUDE)) +
    geom_point(alpha = 0.5, size = 0.5)
```

---
class: inverse
## Your Turn (8 mins)

Connect to the FARS database (using the SQLite database `fars2014`). 
Answer the following questions:

- are there some days of the week where more accidents happen than on others (use variable `DAY_WEEK`)?

- what time of the day do accidents happen (use variable `HOUR`)?

- what is the number of accidents with at least one drunk driver (use variable `DRUNK_DR`)?

---

```{r eval = FALSE}
dayweek <- accidents %>% group_by(DAY_WEEK) %>% tally() 
dayweek %>% collect() %>% 
  ggplot(aes(x = DAY_WEEK, weight=n)) + geom_bar() 

hours <- accidents %>% group_by(HOUR) %>% tally() 
hours %>% filter(HOUR < 25) %>% collect() %>% 
  ggplot(aes(x = HOUR, weight=n)) + geom_bar() 

drunk <- accidents %>% group_by(DRUNK_DR) %>% tally() 
drunk %>% collect() %>% 
  ggplot(aes(x = DRUNK_DR, weight=n)) + geom_bar() 
```

---
class:inverse
## Your Turn

What is a better way of connecting to the sqlite data base without using the deprecated function `src_sqlite`? 

```{r}
src_sqlite
```

---

# Solution to Your Turn

Gut the deprecated function

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), "../data/fars2014")
fars2 <-  dbplyr::src_dbi(con, auto_disconnect = TRUE)
fars2
```

---
class: inverse
## Your Turn (5 mins)

Connect to the  `person` table. Identify drivers (`PER_TYP` == 1, see [fars manual](https://www-fars.nhtsa.dot.gov/Help/Terms.aspx) ) and subset on them.

---

```{r}
person <- tbl(fars, "person")
drivers <- person %>% filter(PER_TYP == 1)
```

---

## Working with multiple tables

- Only rarely single datasets are giving us all the answers we need

- `left_join`, `right_join` work in the same way on databases as on local data frames.

---

## Investigating accidents by gender

We want to join drivers and accidents:

```{r}
driver_acc <- left_join(drivers, accidents)
```

Note that there are a lot of variables with the same name - we actually include all of these variables 

---
class: inverse
## Your Turn (5 min)

- Join drivers and accidents data.

- Tally the number of accidents by day of the week (`DAY_WEEK`), hour of the day (`HOUR`) and gender (`SEX`). Visualize the results!

---

```{r}
gg <- driver_acc %>% group_by(DAY_WEEK, HOUR, SEX) %>% 
  tally() %>% 
  filter(HOUR < 25, SEX < 8) %>% 
  collect() %>% 
  ggplot(aes(x = HOUR, y = n)) + 
    geom_point(aes(colour = factor(SEX))) +
  facet_wrap(~DAY_WEEK, ncol=4) +
  scale_colour_brewer(palette="Set1") +
  theme(legend.position = "bottom")
```

---

```{r, fig.width = 6, fig.height = 4.5}
gg
```

---
class: inverse
## Your Turn

**Creating a SQLite Database**

The data for accidents in 2020 are published on the FARS website in csv form (in 20+ separate files). 

A subset of the tables is available from our github repo: [accident.csv, person.csv, and vehicle.csv]()

Work through the RSQLite intro to create a new database `fars2020` from the three csv files: https://cran.r-project.org/web/packages/RSQLite/vignettes/RSQLite.html

There should be an 'interesting' change in the pattern of accidents :(

---

```{r eval = FALSE}
library(DBI)

fars2020 <- dbConnect(RSQLite::SQLite(), "fars2020.sqlite")

accidents <- read.csv("../data/FARS-2020/accident.csv")
dbWriteTable(fars2020, "accidents", accidents)

person <- read.csv("../data/FARS-2020/person.csv")
dbWriteTable(fars2020, "person", person)

vehicle <- read.csv("../data/FARS-2020/vehicle.csv")
dbWriteTable(fars2020, "vehicle", vehicle)

dbListTables(fars2020)

```

```{r eval = FALSE}
[1] "accidents" "person"    "vehicle"  
```
