---
title: "Stat 585 - elements of the tidyverse"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## tidyverse

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```


`tidyverse` is an 'opinionated' package bundling several other R packages for data science tasks:

```{r}
tidyverse::tidyverse_packages()
```

- share common data representations and API, i.e. work well together

- see https://github.com/hadley/tidyverse for more information

- run the command `tidyverse::tidyverse_update()` to check if/which of your packages need updating


---

## Graphical and numerical summaries


1. graphical summaries - visualizations with `ggplot2`:

    - [RStudio cheatsheet for ggplot2](hhttps://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf)    
    - online documentation: https://ggplot2.tidyverse.org/
    
    - Google group: https://groups.google.com/forum/#!forum/ggplot2
  
2. numerical summaries - elements of `dplyr` and `tidyr`


---

## The pipe operator `%>%`

`f(x) %>% g(y)` is equivalent to `g(f(x), y)`

i.e. the output of one function is used as input to the next function. This function can be the identity

Consequences:

- `x %>% f(y)` is the same as `f(x, y)`

- statements of the form `k(h(g(f(x, y), z), u), v, w)` become
`x %>% 
   f(y) %>% 
   g(z) %>% 
   h(u) %>% 
   k(v, w)`

- read `%>%` as "then do"


---

## dplyr 

There are five primary `dplyr` *verbs*, representing distinct data analysis tasks:

- **Filter**: Select specified rows of a data frame, produce subsets

- **Arrange**: Reorder the rows of a data frame

- **Select**: Select particular columns of a data frame

- **Mutate**: Add new or change existing columns of the data frame (as functions of existing columns)

- **Summarise**: Create collapsed summaries of a data frame

- (**Group By**: Introduce structure to a data frame)

<br>

`dplyr` resources:

- https://dplyr.tidyverse.org/

- R for Data Science book: https://r4ds.had.co.nz/transform.html

---
class: inverse
## Your Turn

The Iowa State Police Department  publishes a Daily Crime Log at https://www.police.iastate.edu/crime-log/

Data for 2017 and most of 2018 are available as file [pd-isu.csv](https://raw.githubusercontent.com/Stat585-at-ISU/materials-2019/master/data/isu-pd.csv) from the course website.

Read the data into your R session and answer the following questions:

- which types of crimes (`Classification`) are typically committed? How many different classifications are there?

- which days are in the top ten for 2018? (Use the `lubridate` package to get to date variables)

- how many times a day are crimes typically reported? 

- what else do you find?


```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
pd <- read_csv("../data/isu-pd.csv")
pd$date <- lubridate::mdy_hm(pd$`Date/Time Reported`)

pd$day <- lubridate::as_date(pd$date)
```

---
class: inverse
## Your turn

The files [isu-football-2017.csv](https://raw.githubusercontent.com/Stat585-at-ISU/materials-2019/master/data/isu-football-2017.csv) and [isu-football-2018.csv](https://raw.githubusercontent.com/Stat585-at-ISU/materials-2019/master/data/isu-football-2018.csv) consist of the Cyclones' football schedule in 2017 and 2018. 

Use the data to create the chart below:

```{r, echo = FALSE, fig.height = 5, fig.width = 8}

schedule <- read.csv("../data/isu-football-2018.csv")
schedule <- schedule %>%
  mutate(
    date = lubridate::ymd("2018-01-01")
    )
lubridate::month(schedule$date) <- as.numeric(factor(schedule$Month, levels=c("Sep", "Oct", "Nov", "Dec"))) + 8
lubridate::mday(schedule$date) <- schedule$Day
schedule$yday <- lubridate::yday(schedule$date)
schedule$year <- lubridate::year(schedule$date)


schedule17 <- read.csv("../data/isu-football-2017.csv")
schedule17 <- schedule17 %>%
  mutate(
    date = lubridate::mdy(Date)
  )
schedule17$yday <- lubridate::yday(schedule17$date)
schedule17$year <- lubridate::year(schedule17$date)

perday <- pd %>% 
  complete(day = full_seq(pd$day, period = 1)) %>% 
  group_by(day) %>%
  summarize(
    n = sum(!is.na(Classification))
  ) 

perday %>%
  mutate(
    year = lubridate::year(day),
    yday = lubridate::yday(day)
    ) %>%
  filter(year <= 2018) %>%
  ggplot(aes(x = yday, y = n)) +
  geom_vline(data = schedule, aes(xintercept = yday, colour = home)) +
  geom_vline(data = schedule17, aes(xintercept = yday, colour = Home)) +
  geom_point(size = .75) +
  facet_grid(year~.) +
  theme_bw() +
  ylab("Number of police reports") +
  xlab("Day of the year")
```

Don't forget to *look at* the chart. What are your main findings?

---

## Tools for working with data

- moving between wide and long forms of data:

    - `pivot_longer`
    
    - `pivot_wider` 
    
    - see https://r4ds.had.co.nz/tidy-data.html


- joining data sets: `left_join`, `anti_join`, see also https://r4ds.had.co.nz/relational-data.html


---
class: inverse
## Your turn

Does the weather affect the number of citations for drug and alcohol-related offenses? 

Use  [Ames_weather_2017-2018.csv](https://raw.githubusercontent.com/Stat585-at-ISU/materials-2019/master/data/Ames_weather_2017-2018.csv) and combine the data with relevant Ames PD data. 

Graphically examine the relationship between temperature and number of citations. You may find it helpful to only consider Thursday - Saturday.

```{r, message = FALSE}
ames_weather <- read_csv("../data/Ames_weather_2017-2018.csv") %>%
  filter(!is.na(time))

# summarize by day to get daily highs and lows
ames_weather_daily <- ames_weather %>%
  mutate(day = as.Date(date)) %>%
  group_by(day) %>%
  summarize(temp_low = min(temp_low), temp_high = max(temp_high))

```

.small[For the curious: [script to get the weather data](https://gist.github.com/srvanderplas/9f4a471e0fb19495935fb2104f3ba9f8)]
