---
title: 'Lab Assignment #2: Exploring the world'
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, echo = TRUE)
```
# Outline

- The package `sf` for drawing maps (and keeping your sanity?)

- Looking under the hood ...



The final deliverable is again a self-contained RMarkdown file. 

---

# Lab organization

1. As last time, there is a github classroom link. Select the invite with your team's number. 

2. Hofmann, don't forget the Zoom breakout rooms. Join the room that matches your team number.

3. Final deliverable: submit a link to your repository in Canvas (This will show me that you are done working on your project).

4. Create a folder `data` in the repository. Assume that all of the data files used are inside this folder, similar to the code used in this Rmarkdown file. Upload the gadm data, but not the middle earth data. 



---

# First item: talk!

Just as with lab #1, you are asked to create one file (README.Rmd) with your results. 

All team members are supposed to contribute ... last time you might have realized that it is a pretty big mess if everybody writes into the README file all at the same time.

This time: figure out a plan before you start working.

---

# The package `sf`

A lot of government agencies use shapefiles to publish spatial information, such as states, counties, congressional districts, exact locations of state and national parks, etc. 

The `sf` package is a new(ish) development that is likely going to replace packages `sp` and `maptools` for working with spatial data (https://r-spatial.github.io/sf/articles/sf1.html).

**Install the `sf` package from CRAN. You might have to install additional dependencies.**


<!--You might to also have to install `rgdal` (R package) and the gdal framework (https://www.gdal.org/).-->

---

# Fun with Middle Earth

Not just the government, dedicated fan bases alike provide information in shapefiles. 
The ME-DEM team recreates Tolkien's Middle Earth in shapefile format at https://github.com/jvangeld/ME-GIS

Download the ME-GIS repository, put it into the data folder, but **do not push this folder to your repo** (We have to respect intellectual property). 

---

```{r}
library(ggplot2)
library(sf)

rivers <- read_sf("data/ME-GIS/Rivers.shp")
roads <- read_sf("data/ME-GIS/Roads.shp")

p <- ggplot() +
  geom_sf(data = read_sf("data/ME-GIS/Coastline2.shp"), 
          colour="grey10", fill="grey90") +
  geom_sf(data = rivers %>% dplyr::filter(TYPE=="Stream"), 
          colour="steelblue", size=0.3) +
  geom_sf(data = rivers %>% dplyr::filter(TYPE=="River, Unknown Length"), 
          colour="steelblue", size=0.1) +
 geom_sf(data = roads %>% dplyr::filter(TYPE == "PRIMARY"),
         size = 0.7, colour="#663300") +
 geom_sf(data = roads %>% dplyr::filter(TYPE == "SECONDARY"),
         size = 0.5, colour="#663300") +
 geom_sf(data = roads %>% dplyr::filter(TYPE == "TERTIARY"),
         size = 0.3, colour="#663300") +
  geom_sf(data = read_sf("data/ME-GIS/Cities.shp"), 
          colour ="#336699", size = 3.5) +
  theme_bw() 

```

---

Now let's plot Middle Earth:
```{r}
p
```

---
class: inverse
# Todo for you: more map elements

- Add labels for the cities (see  `geom_sf_text`) or maybe some regions?
Where's the Shire? Where's Mordor?

Getting Fancy:

- To appreciate how far Frodo and Sam had to travel, include a scale with `annotation_scale` (`ggspatial`).
- Also show North in a more map typical way with `annotation_north_arrow` (`ggspatial`)

<br>

**Include the code to produce the final map in your  RMarkdown file.**
        
---

# Looking under the hood ...

The main objective of this part is to write code that allows us to read in data from a shapefile and access the data in a way that allows us to use it in the `ggplot2` framework, i.e essentially we want to figure out what goes on behind the scenes after using `sf_read`.

We will be using the australia shapefile for a concrete example and then try to bundle  general ideas into a function. 

---

# Quick review

A map in ggplot2 consists a couple of variables to make things work:

```{r}
library(ggplot2)
library(maps)
states <- map_data("state")
head(states)
```

- x and y variable for longitude and latitude. 
- `group` identifies a polygon part. 
- `order` specifies in which order the points in a polygon connect. 
- `region` usually allows us to connect information from the outside to the map. 
- `subregion` has no function for the states map, but takes on functionality for other maps

---

```{r, message=FALSE}
library(dplyr)
michigan <- 
  states %>% 
  filter(region=="michigan") %>%
  group_by(group) %>% 
  summarize(
    region = region[1],
    n = n())

michigan
```



---

```{r, fig.width=5, fig.height=3, fig.align='center'}
ggplot(states, aes(x=long, y=lat, group=group)) + geom_polygon()
```




---

# Australia shapefile

For a concrete example, we will load in the data from the  shapefile for Australia

```{r}
ozbig <- read_sf("data/gadm36_AUS_shp/gadm36_AUS_1.shp")
```

Generally shape files are very big - we need to thin the number of points used. That's what the following lines do. Make sure to run those, or you will likely crash your R session.
```{r}
oz_st <- maptools::thinnedSpatialPoly(
  as(ozbig, "Spatial"), tolerance = 0.1, 
  minarea = 0.001, topologyPreserve = TRUE)
oz <- st_as_sf(oz_st)
```

---

# sf object

```{r}
oz
```

---

`oz$geometry` is a list variable

```{r}
is.list(oz$geometry)
```

Each element in this list consists of another list: 

```{r}
str(oz$geometry[[1]])
```

---

# Inner workings

Now we work our way into the inner structure of the `geometry` object.

Each element of the form `oz$geometry[[i]][[j]]` is another list with one element: a matrix with two columns

```{r}
head(oz$geometry[[1]][[3]][[1]])
```

These values give x and y coordinates of points of a polygon. 

---
class: inverse
# Your Task

Write a helper function to turn one of these matrices into a single data frame with variables `long`, `lat`, `group`, `order`.
Use `purrr` functionality to create a dataframe `ozplus` from the `geometry` variable. Plot the result with the command:

```
ozplus %>% ggplot(aes(x = long, y = lat, group = group)) + geom_polygon()
```

Include all the code necessary to read the shape file and convert to the data frame plotted in your Rmarkdown file. 
Include a plot of the result as well. 

---

# Testing your code

How reliable is your code? - does it work on anything but the `oz` example? 

There's only one way to answer that question! Try it out!

In the next step, you are asked to get a shapefile of a country of your choice and use your code on the new shape file. 

---
class: inverse
# GADM

GADM is a service "to provide boundaries without borders" 

- Go to the website at https://gadm.org/download_country_v3.html and download a shape file of a country of your choice.

- Unzip the document and move the resulting folder into your working directory 

- Read the shapefile (there are usually shapefiles at different administrative levels - pick one). Thin the shapefile (or you might crash your R session; shapefiles can be big)
- Export the polygons and plot in `ggplot2`.

Include the code and the result in your README.Rmd file. 

<br>

**Upload a link to your repository to Canvas. Only one submission per team is required. 
Finishing touches can be made until Monday, Feb 21, 11:59 pm.**
