---
title: 'Lab Assignment #2: Couple of extra hints'
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, echo = TRUE)
```
# Couple of extra hints

- GADM shapefiles and Australia

- matrices


---

# GADM shapefiles

Link: https://gadm.org/download_country_v3.html provides access to a lot of different countries

Various formats for shapefiles are available

  - zipped folder with files: Shapefiles 
  - R binary (rds) for each level 0, 1, 2
  
Levels describe country (0), states or territories (1), and counties or other districts (2)
  
  
---

# Australia - download zipped folder

Download file from a path, unzip it and view the contents: 
```{r}
# download
zipped <- "https://biogeo.ucdavis.edu/data/gadm3.6/shp/gadm36_AUS_shp.zip"
download.file(zipped, destfile="data/gadm36_AUS.zip", mode="wb")

# unzip
unzip("data/gadm36_AUS.zip", exdir ="data/gadm36_AUS/")

# view shapefiles (extension .shp)
dir("data/gadm36_AUS/", pattern = "shp")
```

---

# Read shape files for all levels 


```{r}
library(sf)
oz_0 <- read_sf("data/gadm36_AUS/gadm36_AUS_0.shp")
oz_1 <- read_sf("data/gadm36_AUS/gadm36_AUS_1.shp")
oz_2 <- read_sf("data/gadm36_AUS/gadm36_AUS_2.shp")
```

Shapefiles can be quite big -  look at their size first before plotting

```{r message = FALSE}
library(maptools)
object.size(as(oz_0, "Spatial"))
object.size(as(oz_1, "Spatial"))
object.size(as(oz_2, "Spatial"))

```
---

# Thinning spatial structure

Thinning algorithm (Douglas-Pecker algorithm) reduces the number of nodes in a polygon (works similar to smoothing by removing some of the small scale structure).

Too much thinning will result in unrecognizable objects.


```{r}
thin <- function(x, tolerance = 0.1, minarea = 0.01) {
  xx <- maptools::thinnedSpatialPoly(
    as(x$geometry, "Spatial"), 
    tolerance, minarea , topologyPreserve = TRUE)
  x$geometry <- st_as_sf(xx)$geometry
  x
}

object.size(oz_0)
object.size(thin(oz_0))
```

---

# Plot the shape files 

```{r fig.height=3}
library(ggplot2)
library(patchwork)
p0 <- ggplot() + 
  geom_sf(data = thin(oz_0), colour = "red") + ggtitle("Level 0")
p1 <- ggplot() +  
  geom_sf(data = thin(oz_1), colour = "blue") + ggtitle("Level 1")
p2 <- ggplot() + 
  geom_sf(data = thin(oz_2), colour = "purple") + ggtitle("Level 2")
p0+p1+p2
```

---

# RDS files

Work extremey similar to the shapefiles - download rds file at level of choice, thin, and plot: 
```{r}
# download
rds_0 <- "https://biogeo.ucdavis.edu/data/gadm3.6/Rsp/gadm36_AUS_0_sp.rds"
download.file(rds_0, destfile="data/rds_0.rds", mode="wb")
ozrds_0 <- readRDS("data/rds_0.rds")

class(ozrds_0)
```

```{r}
ggplot() + 
  geom_sf(data = thin(st_as_sf(ozrds_0)))
```

---

# Inner workings

Polygons are made up of sets of points containing geographic latitude and longitude, individual objects (such as e.g. Michigan) can be made up of multiple polygons.

```{r message= FALSE}
library(dplyr)
oz_1_sm <- thin(oz_1) # switch to thinned version
oz_1_sm <- oz_1_sm %>% mutate(
  numPoly = geometry %>% purrr::map_dbl(.f = length) 
)
head(oz_1_sm$numPoly)
```

---

# Setting up a Before picture

```{r}
ggplot() + geom_sf(data = oz_1_sm[1,])
```

---

```{r}
length(oz_1_sm$geometry[[1]])
oz_1_sm$geometry[[1]][[1]] # each element is a list itself
```
---


```{r}
oz_1_sm <- oz_1_sm %>% mutate(
  data = 1:length(geometry) %>% purrr::map(.f = function(i) {
    g <- geometry[[i]]
    res <- g %>% purrr::map_df(.id="i", .f = function(gelement) {
      as.data.frame(gelement[[1]]) 
    })
    
    res$group <- paste0(i,"-",res$i)
    res
  })
)
```
---

# Compare to Before pic

```{r}
oz_1_sm$data[[1]] %>%
  ggplot(aes(x =x, y = y, group = group)) + geom_polygon()
```
---

Still to do: unnest the data, rename variables, introduce order variable
