---
title: 'Lab Assignment #3 - Recap'
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, middle, center
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, echo = TRUE)
```
# A systems engineer walks into a bar ...

---
class: inverse
# ... 

- he orders a beer

- he orders 9,999,999 beers

- he orders -1 beers

- he orders C5F63A beers

- he orders a lizard

---
class: inverse
# the first real customer enters the bar

... and asks where the bathroom is.


--

<br> <br> <br>

The bar catches fires, explodes and kills everyone ????????????

<br> <br> <br>

[Thanks Reddit](https://www.reddit.com/r/Jokes/comments/a4a4jc/a_software_qa_engineer_walks_into_a_bar/)


---

# What I will do to your package

1. download repo from github 

2. install and run checks 

3. run an example:

```{r eval = FALSE}
library(tidyverse)
library(pkg_name)
oz <- team_X("path to some shp file")
oz %>% 
  ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = NAME_1))
```

4. run your examples 

5. knit your README.Rmd file and read through it

6. run some additional examples

---

## Why do we do this to you?

- expectation: all of the checks work without problems.

- If the above is not true, use our previous method: 
 when you encounter a problem, make a note of it and put it into the Readme file. 
 (comes with a deduction of points)

- this lab is a technical test run for the final project

- new due date: tonight, 11:59pm


---

## Final project

it looks like you are all set for the final project

Next deliverable: Stage 2 (by Apr 14, not graded): Functional Package

---

# How do things look like?

```
── R CMD check results ──────────────────────────────────────── Lab3package 0.0.0.9000 ────
Duration: 25s

> checking R code for possible problems ... NOTE
  plot_shapefile: no visible binding for global variable ‘lat’
  team_1 : get_df: no visible binding for global variable ‘.’
  team_1: no visible binding for global variable ‘.’
  team_3: no visible binding for global variable ‘.’
  team_5 : func.oz: no visible binding for global variable ‘geometry’
  ...
  Undefined global functions or variables:
    . ENGTYPE_1 NAME_0 NAME_1 data district geometry group lat long
  Consider adding
    importFrom("utils", "data")
  to your NAMESPACE file.

0 errors ✓ | 0 warnings ✓ | 1 note x
```
---

# Generally ....


- Do **NOT** use any global file paths, i.e. file names of the form: `C:/Downloads/project/lab-2-lab2-team2/data/gadm36_AUS_shp/gadm36_AUS_1.shp`


Citing [Jenny Bryan](https://github.com/jennybc/here_here#admitting-you-have-a-problem):

> If the first line of your #rstats script is setwd("C:\Users\jenny\path\that\only\I\have"), I will come into your lab and SET YOUR COMPUTER ON FIRE.

More [rage tweets](https://twitter.com/JennyBryan/status/673240213313851393)

<br><br>

- Do not use spaces in file names:

```
> checking for portable file names ... WARNING
  Found the following file with a non-portable file name:
    R/team 4.R
  These are not fully portable file names.
```  


---

# Dependencies

```
> checking dependencies in R code ... WARNING
  'library' or 'require' calls not declared from:
    ‘dplyr’ ‘ggplot2’ ‘purrr’ ‘sf’
  'library' or 'require' calls in package code:
    ‘dplyr’ ‘ggplot2’ ‘purrr’ ‘sf’
    Please use :: or requireNamespace() instead.
    See section 'Suggested packages' in the 'Writing R Extensions' manual.
```

Do **NOT** put calls to `library()` into any function code. 

Function `fff` from package `xyz` are made available by 

1. Placing  package `xyz` in the DESCRIPTION file (under Imports)
2. using the roxygen tag `#' @importFrom xyz fff` where the function is used

---

# No visible binding ...

```
> checking R code for possible problems ... NOTE
  plot_shapefile: no visible binding for global variable ‘lat’
  team_1 : get_df: no visible binding for global variable ‘.’
```

For `lat`: simple solution: include the line in the function (before `lat` is used):

```
lat <- NULL 
```

Other solutions and some explanations for the  [no visible binding ](https://www.r-bloggers.com/2019/08/no-visible-binding-for-global-variable/) problems

--

Use `.data` instead of `.`, e.g.  

More on [use .data not .](https://dplyr.tidyverse.org/articles/programming.html)

---

# Make use of file .Rbuildignore

```
> checking for hidden files and directories ... NOTE
  Found the following hidden files and directories:
    .github/.keep
    .github
  These were most likely included in error. See section ‘Package
  structure’ in the ‘Writing R Extensions’ manual.

> checking top-level files ... NOTE
  Non-standard files/directories found at top level:
    ‘README.Rmd’ ‘lab2-scripts’
```

Expand file .Rbuildignore and include README files, github content. Delete lab2-scripts?

```
^.*\.Rproj$
^\.Rproj\.user$
^README.*  # ignore all files that start with README
^.github$
```
---

# Data examples

```
> team_2("data/gadm40_AUS_shp/gadm40_AUS_1.shp",0.01)
  Error: Cannot open "data/gadm40_AUS_shp/gadm40_AUS_1.shp"; The file doesn't seem to exist.
  Execution halted
  
> checking package subdirectories ... WARNING
  Subdirectory ‘data’ contains no data sets.

> checking contents of ‘data’ directory ... WARNING
  Files not of a type allowed in a ‘data’ directory:
    ‘gadm40_AUS_shp’
  Please use e.g. ‘inst/extdata’ for non-R data files  
```

Where to place the data for your example?

For your example you will need a set of five files: cpg, dbf, prj, shp, and shx

Place non-R data in the folder `inst`

During the installation all elements of `inst` are placed in a special location on your machine.

For example, to find `inst/extdata/mydata.csv`, you'd call `system.file("extdata", "mydata.csv", package = "mypackage")`

---

# Example of an example

Assume you have the australia files in a folder called `extdata` in folder `inst`

The example then should be:
```{r eval=FALSE}
ozfilepath <- system.file("extdata/gadm36_AUS_1.shp", ".shp", package = "mypackage")
ozdata <- team_X(ozfilepath, tolerance=0.1)
```

---

# Another solution to incorporate data

save spatial data frame as rda object, i.e. for thinned Australia shapes `oz`:

```
usethis::use_data(oz)
```

Then use `data(oz)` in your examples.

Needs to be documented in file data.r
