---
title: "Stat 585 - Iterations"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, "tweaks.css"]
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```
## Iterations

- Run the same block of code multiple times:
```{r iterations1, eval = F}
for (i in players) {
  #####################
  # Block of commands #
  #####################
  
  # Output
}
```

---

## Iterations

- Run the same block of code multiple times:
```{r iterations2, eval = F}
for (i in players) {
  #####################
  # Block of commands #
  #####################
  
  print(avg)
}
```


---

## Iterations

- Run the same block of code multiple times:
```{r iterations3, eval = F}
for (i in players) {
  
  player <- subset(baseball, playerID == i)
  avg <- sum(H/AB, na.rm = T)
  
  print(avg)
}
```

--

This is a loop, or iteration

---

## Iterations and R

- R is vector-based - many tasks don't require loops at all

- `for` loops are useful for debugging purposes

There are several alternatives to `for` loops in R. We will look into some alternatives next time:

- The `apply` series of functions (`lapply`, `sapply`, `mapply`, `tapply`) can be used to "loop" in R without an explicit `for` loop

- `purrr` is a tidyverse package used for functional programming
    - tidy version of `apply`: `map`
    - more consistent API


---
class: inverse
## Your Turn

[FizzBuzz](http://www.dr-mikes-math-games-for-kids.com/fizz-buzz.html) is a common game to teach children multiplication tables

Write a function that, for any number between lower bound $a$ and upper bound $b$, outputs the correct response to FizzBuzz:

- if the number is divisible by 3, the program returns "fizz" instead of the number
- if the number is divisible by 7, the program returns "buzz" instead of the number
- if the number is divisible by 3 and 7, the program returns "fizzbuzz" instead of the number
- in any other case return the number unchanged

The modulus operator, `%%` will be useful.

1. Write a fizzbuzz function using a `for` loop

2. Can you make your function work without using a loop?

---
class: middle, center

# no peeking before trying!

---

it's usually a good idea to break down tasks into very simple steps

```{r}
fizzbuzz1 <- function(x) {
  # we assume x is a single number
  div3 <- (x %% 3) == 0
  div7 <- (x %% 7) == 0
  
  if (div3 & div7) return("fizzbuzz")
  if (div3) return("fizz")
  if (div7) return("buzz")
  return(x)
}

fizzbuzz1(5)
fizzbuzz1(21)
```

---

Now wrap `fizzbuzz1` into a bigger function

```{r}
fizzbuzz <- function(a, b) {
  # we assume a and b are integer values
  
  for (i in a:b) fizzbuzz1(i)
}

fizzbuzz(1,10)
```

Why isn't anything happening?

---

We need to make some space to save values

```{r}
fizzbuzz <- function(a, b) {
  # we assume a and b are integer values
  
  idx <- a:b
  result <- rep(NA, length(idx))
  for (i in 1:length(idx)) 
    result[i] <- fizzbuzz1(idx[i])
  
  result
}

fizzbuzz(1,10)
```

---

# Solution without a loop

```{r}
fizzbuzz_no_iter <- function(a, b) {
  lower <- min(a, b)
  upper <- max(a, b)
  
  if (lower == upper) return(fizzbuzz1(lower))
  
  return( c(fizzbuzz1(lower), fizzbuzz_no_iter(lower+1, upper)))
}

fizzbuzz_no_iter(1, 10)
```

---

# ... and another solution without a loop

```{r}
fizzbuzz_vector <- function(a, b) {
  # create a vector with numbers from a to b
  x <- seq.int(from = min(a,b), to = max(a,b))
  res <- x
  
  res[x %% 3 == 0] <- "fizz"
  res[x %% 7 == 0] <- "buzz"
  res[(x %% 3 == 0) & (x %% 7 == 0)] <- "fizzbuzz"
  
  res
}

fizzbuzz_vector(1,10)
```
---

# Still to do later  ...

Which of the functions is 'best'?

Best as in: fastest, most readable, most maintainable

... we will discuss profiling code and other aspects of good coding practices later


---

# Functions of functions

- The output of a function can also be a function

- The fizzbuzz game can be played with any two integer values besides 3 and 7

- We could therefore make a function `fbgen (m, n)` that makes a fizzbuzz function for integers `m` and `n`

---

```{r}
fbgen <- function(m, n) {
  # couple of error checks
  stopifnot(is.integer(m), is.integer(n), 
            length(m) == 1, length(n) == 1,
            m != n)
  
  res <- function(x) {
    stopifnot(is.numeric(x))
    
    out <- x 
    out[x %% m == 0] <- "fizz"
    out[x %% n == 0] <- "buzz"
    out[x %% (m*n) == 0] <- "fizzbuzz"
    
    return(out)
  }
  
  return(res)
}

```

---


```{r}
fb23 <- fbgen(2L,3L) # the L turns 2 and 3 into integers
fb23

fb23(1:10)
```

---
class: middle, center
# ... take another approach to iterations
