---
title: "Stat 585 - How do we write good code?"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, default-fonts, "tweaks.css"]
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

## Signs of inexperienced programmers


- Violation of the language model
    - e.g. R is vector oriented language, i.e.  ```for``` loops should be used sparingly, 
    - e.g. use of ```rbind``` (it is *known* to be very slow)

- Insufficient knowledge of the language. We're all guilty of that at different levels 

- Reading other people's code helps improve our language knowledge!

---

## Removing for loops

`x` is vector with numbers 1,2,3,4

Objective: re-code these into weather events: "Sun", "Rain", "Snow", "Hail"

```{r}
x <- sample.int(4, size=100, replace = TRUE)
y <- vector("character", length=length(x))
for (i in 1:length(x)) {
   y[i] <- switch (x[i], "Sun", "Rain", "Snow", "Hail", default=NA)
}

table(y)
```

---

## Removing for loops (cont'd)

```{r}
# the switch statement on the previous slide is equivalent to 
# the following lines:
for (i in 1:length(x)) {
  if (x[i] == 1) y[i] <- "Sun"
  if (x[i] == 2) y[i] <- "Rain"
  if (x[i] == 3) y[i] <- "Snow"
  if (x[i] == 4) y[i] <- "Hail"
}

# using R's vector system we can reduce this to a single line:
y <- c("Sun", "Rain", "Snow", "Hail")[x]
```

---

## Removing for loops (cont'd)

All of the previous solutions doctor around symptoms

We are still breaking the language model :)

--

```{r}
y <- factor(x, labels =  c("Sun", "Rain", "Snow", "Hail"))

summary(y)
```

---

## Spotting bad code

- **Ken Thompson:** <br> keep the number of indentations down, high number of local variables is suspicious

- **Hadley Wickham:** <br> functions with more than ten lines are suspicious, as is highly repetitive code

- **Generally Good Practices:** <br> structure code using indentation and spacing, write comments, use clear naming convention

---

## Profiling

Code can be profiled at different levels:

1. two (or more) alternatives of code with the same objectives are timed to determined the one which is the fastest (`system.time()` and `proc.time()`)

2. one piece of code is analyzed to find weak/slow spots (e.g. `profvis` )

---

## Profiling process

- Collect information on code performance:

- memory usage

- frequency and duration of function calls

- Goal: optimization of code

---

## `system.time()` or `proc.time()`

- Probably simplest of all measures

- Evaluates the time spent in the whole expression

- doesn't give any clues about what is done at a lower level <br><br>


---

## (Mini) code blocks


```{r message=FALSE}
ptm <- proc.time()
t1 <- read.csv(here::here("data/temps.csv"))
proc.time() - ptm

ptm <- proc.time()
t2 <- readr::read_csv(here::here("data/temps.csv"), progress = FALSE)
proc.time() - ptm
```

---

## (Mini) code blocks

```{r message=FALSE}
ptm <- proc.time()
t3 <- readRDS(here::here("data/temps.rds"))
proc.time() - ptm

ptm <- proc.time()
load(here::here("data/temps.Rda"))
proc.time() - ptm
```


---
class: inverse
## Your turn: which option is fastest?

```{r message=FALSE}
x <- sample.int(4, 100, replace=TRUE)
y <- vector("character", length=length(x))

# Option #1
for (i in 1:length(x)) {
   y[i] <- switch (x[i], "Sun", "Rain", "Snow", "Hail", default=NA)
}

# Option #2
for (i in 1:length(x)) {
  if (x[i] == 1) y[i] <- "Sun"
  if (x[i] == 2) y[i] <- "Rain"
  if (x[i] == 3) y[i] <- "Snow"
  if (x[i] == 4) y[i] <- "Hail"
}

# Option #3
y <- c("Sun", "Rain", "Snow", "Hail")[x]

# Option #4
y <- factor(x)
levels(y) <- c("Sun", "Rain", "Snow", "Hail")

```
---

## Your Turn Result

Based on 500 evaluations of samples of size 5000: 

```{r message=FALSE, warning=FALSE, echo = FALSE, fig.height = 6, fig.width = 7}
library(tidyverse)
set.seed(20190310) # remember the rules of reproducibility!

results <- 1:500 %>% purrr::map_df(.f = function(x) {
  x <- sample.int(4, 5000, replace=TRUE)
  y <- vector("character", length=length(x))
  
  ptmbase <- proc.time()
  # Option #1
  for (i in 1:length(x)) {
    y[i] <- switch (x[i], "Sun", "Rain", "Snow", "Hail", default=NA)
  }
  ptm1 <- proc.time()
  
  # Option #2
  for (i in 1:length(x)) {
    if (x[i] == 1) y[i] <- "Sun"
    if (x[i] == 2) y[i] <- "Rain"
    if (x[i] == 3) y[i] <- "Snow"
    if (x[i] == 4) y[i] <- "Hail"
  }
  
  ptm2 <- proc.time()
  
  # Option #3
  y <- c("Sun", "Rain", "Snow", "Hail")[x]
  
  ptm3 <- proc.time()
  
  # Option #4
  y <- factor(x, labels=c("Sun", "Rain", "Snow", "Hail"))

  
  ptm4 <- proc.time()
  data.frame(value=c((ptm1-ptmbase)["elapsed"], (ptm2-ptm1)["elapsed"], (ptm3-ptm2)["elapsed"], (ptm4-ptm3)["elapsed"]), Option=1:4)
})

results %>% 
  ggplot(aes(x = Option, y = value, fill= factor(Option))) +geom_boxplot() +
  ylab("Elapsed time") + ylim(c(0, 0.03))
```
---

## Profiling Code

Option 3 is  fastest on average. But why?

`profvis` package:

- wrapper around function `Rprof`

- call stack is written every 0.02 seconds (interval) into a temporary file

- output file is processed to produce a summary of the usage

- time measurements varies by platform: on Unix-alike time is CPU time of the R process (excludes time waiting for input)

- Note: profiling takes time, too; once the timer goes off, the information is not recorded until the next timing click (probably in the range 1–10 msecs). Can’t go below that.


---

## `profvis` example

```{r eval=FALSE}
library(profvis)

p <- profvis({			
  data(diamonds, package = "ggplot2")			
  plot(price ~ carat, data = diamonds)			
  m <- lm(price ~ carat, data = diamonds)			
  abline(m, col = "red")			
})
```

`profvis` creates interactive response with code and flame plot.

Copy and paste the code into your console to run it.

---
class: inverse
## Your Turn

Remember the fizzing and the buzzing we did at the beginning of the semester? 

The file [`fizzbuzz-ing.R`](../code/fizzbuzz-ing.R) contains three different implementations of the `fizzbuzz` function.

Test which function is the fastest on average.

Where is the time spent during processing?

